Place of death in NRS COVID-19 deaths

### 2021-08-17 covid_nrs_place_of_death_visualisations.Rmd
updated 2020 & 2021 on same graph plot, and added plots showing proportion of home deaths by geography compared to scottish average

@@

```{r}
(figure_deaths_at_home_2020_and_2021 <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # geom_line(aes(y = deaths, colour = deaths_type, alpha = year)) +
  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number), show.legend = FALSE) +
  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.8), guide = NULL) +
  annotate(geom = "text", x = 18, y = 550, label = "2020", alpha = 0.6) +
  annotate(geom = "text", x = 4, y = 530, label = "2021") +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 10, y = 100, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 10, xend = 14, y = 110, yend = 250, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 50, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 60, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)

@@

## HB

```{r}
# (figure_proportion_home_deaths_over_time <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "hb") %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(size = 1) +
  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  # facet_wrap(~ ref_area, scales = "free_x") +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_x_continuous(breaks = 2015:2021) +
  # scale_y_continuous(breaks = seq(0.1, 0.6, by = 0.1), label = function(x) scales::percent(x, accuracy = 1)) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  # theme_minimal() +
  theme(
    legend.position = "top", 
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80")
    ) +
  labs(
    x = "Year", y= "Proportion of yearly deaths (%)",
    colour = "",
    subtitle = paste0("* ",only_2021_text),
    caption = paste0(c(source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
# )

>>>>

```{r}
(figure_deaths_at_home_2020_and_2021 <-
  merged_deaths_overall %>%
  filter(sex == "all" & age == "all") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  select(-c(date_w_c, week_number_run_over, ref_period, ref_area)) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number, group = deaths_type), show.legend = FALSE) +
  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.3) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 600) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
  annotate(geom = "text", x = 18, y = 550, label = "2020", alpha = 0.4) +
  annotate(geom = "text", x = 4, y = 530, label = "2021") +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 10, y = 100, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 10, xend = 14, y = 110, yend = 250, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 50, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 60, yend = 270, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)

@@

@@

## all places of death, HB

```{r}
(figure_proportion_of_deaths_by_place_by_hb <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "hb") %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point(aes(size = place_of_death)) +
  geom_line(aes(size = place_of_death)) +
  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = 2015:2021) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80")
    ) +
  labs(
    x = NULL, y= "Proportion of yearly deaths (%)",
    subtitle = paste0(c(paste0("* ",only_2021_text),"Health boards sorted by 2020 population estimate."),collapse="\n"),
    caption = paste0(c(source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2, title = NULL), 
         size = guide_legend(nrow=2, title = NULL)) +
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_by_place_by_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_by_place_by_hb, width = 12, height = 10, units = "in", dpi = 300)
```

## all places of death, LA

```{r}
(figure_proportion_of_deaths_by_place_by_la <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "la") %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange hb/la by population size
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point(aes(size = place_of_death)) +
  geom_line(aes(size = place_of_death)) +
  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE, ncol = 4) +
  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = 2015:2021) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80")
    ) +
  labs(
    x = NULL, y= "Proportion of yearly deaths (%)",
    subtitle = paste0(c(paste0("* ",only_2021_text),"Local authorities sorted by 2020 population estimate."),collapse="\n"),
    caption = paste0(c(source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2, title = NULL), 
         size = guide_legend(nrow=2, title = NULL)) +
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_by_place_by_la"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_by_place_by_la, width = 12, height = 14, units = "in", dpi = 300)
```

## Home deaths + Scottish average, HB

```{r}
(figure_proportion_of_deaths_at_home_by_hb <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "hb") %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
  geom_point() +
  geom_line() +
  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE) +
  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = 2015:2021) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80")
    ) +
  labs(
    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
    linetype = NULL,
    subtitle = paste0(c(paste0("* ",only_2021_text),"Health boards sorted by 2020 population estimate."),collapse="\n"),
    caption = paste0(c(source_nrs_text),collapse="\n")
    ) +
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_at_home_by_hb"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_at_home_by_hb, width = 12, height = 10, units = "in", dpi = 300)
```

## Home deaths + Scottish average, LA

```{r}
(figure_proportion_of_deaths_at_home_by_la <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "la") %>%
  filter(place_of_death == "Home & other non-institution") %>%
  left_join(merged_proportions_of_deaths_by_place %>% select(year, place_of_death, proportion_scotland = proportion_of_total), by = c("year","place_of_death")) %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(ref_area = factor(x = ref_area, levels = order_la_by_population)) %>%  # arrange la/la by population size
  pivot_longer(cols = c(proportion_of_total, proportion_scotland), names_to = "measure", values_to = "proportion_of_total") %>%
  mutate(measure = if_else(measure == "proportion_of_total", true = "Health board", false = "Scotland")) %>%
  ggplot(., aes(x=year, y=proportion_of_total, linetype = measure)) +
  geom_point() +
  geom_line() +
  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  facet_rep_wrap(~ ref_area, repeat.tick.labels = TRUE, ncol = 4) +
  scale_size_manual(values = c(1,1.8,1,1), name = "Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  scale_x_continuous(breaks = 2015:2021) +
  scale_y_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="C", name="Place of death", labels = order_of_place_of_death_levels[c(1,3,2,4)]) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80")
    ) +
  labs(
    x = NULL, y= "Proportion of yearly deaths that happened at home or other non-institution (%)",
    linetype = NULL,
    subtitle = paste0(c(paste0("* ",only_2021_text),"Local authorities sorted by 2020 population estimate."),collapse="\n"),
    caption = paste0(c(source_nrs_text),collapse="\n")
    ) +
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_proportion_of_deaths_at_home_by_la"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_of_deaths_at_home_by_la, width = 12, height = 14, units = "in", dpi = 300)
```

## HB: compact plot of home deaths only

```{r}

# (figure_proportion_of_deaths_by_place_by_hb <-
  proportions_of_deaths_by_place_by_geography %>%
  filter(geography == "hb") %>%
  filter(place_of_death == "Home & other non-institution") %>%
  mutate(ref_area = factor(x = ref_area, levels = order_hb_by_population)) %>%  # arrange hb/la by population size
  mutate(year = factor(x = year, levels = 2015:2021)) %>%  # arrange hb/la by population size
  ggplot(., aes(x=proportion_of_total, y=ref_area, colour=year, shape=year)) +
  geom_point(size = 2) +
  scale_shape_manual(values = c(rep(4, times = 5), rep(16, times = 2))) +
  scale_x_continuous(label = function(x) scales::percent(x, accuracy = 1)) +
  scale_colour_viridis_d(option="D", direction = -1) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 12),
    plot.caption = element_text(size = 10, colour = "gray60"),
    panel.border = element_rect(fill = NA, colour = "gray80"),  # draw a transparent rectangle as the border
    strip.background = element_rect(fill = NA, colour = "gray80")
    ) +
  labs(
    x = NULL, y= "Proportion of yearly deaths (%)",
    colour = NULL,
    subtitle = paste0("* ",only_2021_text),
    caption = paste0(c("Health boards sorted by 2020 population estimate",source_nrs_text),collapse="\n")
    ) +
  # guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
# )
```
